---
  title: "Data Retrieval for Marine Species Distribution Modelling"
author: "Rutendo Musimwa, Willem Boone and Johannes Nowe "
date: "2024-09-24"
output:
  html_document:
  toc: true
toc_float:
  collapsed: false
toc_depth: 2
number_sections: false
theme: default
highlight: default
df_print: paged
css: bootstrap.css
self_contained: false
bibliography: references.bib
csl: apa.csl
fig_width: 10
fig_height: 6
editor_options: 
  markdown: 
    wrap: 72
  wrap: 72
---

```{=html}
<style>
.htmlnb-widget {
  margin: auto;
}
</style>
```
# Load packages and setup environment

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/home/HSM/")
options(paged.print = FALSE)

# load packages
pckgs <- c("arrow", "tidyverse", "doParallel", "rasterVis", "mapview", 
           "ENMeval", "dynamicSDM", "gridExtra", "raster", "spThin",
           "BiocManager","Rarr","worrms","stars","foreach","terra")

invisible(lapply(pckgs, library, character.only = TRUE))
rm(pckgs)
```

# 1. OCCURRENCE DATA

This document demonstrates how to retrieve occurrence data for marine species from the EDITO platform.

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Step 1: Establish data lake connection (S3F)

use `S3FileSystem` from package `arrow` to connect to the data lake.

```{r}
data_lake <- S3FileSystem$create(anonymous = TRUE, scheme = "https", endpoint_override = "s3.waw3-1.cloudferro.com")
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Step 2: Access EurOBIS occurrence data stored in the parquet file.

The EurOBIS data is stored at following location in the data lake:

`emodnet/biology/eurobis_occurence_data/eurobisgeoparquet/eurobis_no_partition_sorted.parquet`

Using this address, you can open the dataset using `arrow::open_dataset`

```{r}
path_to_eurobis = "emodnet/biology/eurobis_occurence_data/eurobisgeoparquet/eurobis_no_partition_sorted.parquet"

eurobis <- arrow::open_dataset(data_lake$path(path_to_eurobis))
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Step 3: Filter the occurrences

You can filter the data set on following criteria:

-   `aphiaidaccepted`: int
-   `latitude` & `longitude`: int or float
-   `Date`: string formatted date `YYYY-MM-DD`

Some example Aphia IDs

| Species           | Aphia ID |
|-------------------|----------|
| Atlantic mackerel | 127023   |
| Atlantic herring  | 126417   |
| European seabass  | 126975   |

More Aphia IDs can be found using the `worrms` package to query the worms database. Searching on scientific name or common name:

```{r}
# Searching on scientific name
worrms::wm_name2id("Clupea harengus")

# Searching on common name
worrms::wm_records_common("Atlantic herring")|> dplyr::select("AphiaID","scientificname")
```

Define your parameters here: For example for herring:

```{r}
aphia_ID = 126417
sel_longitude = c(0, 1)
sel_latitude = c(50, 51)
start_date = "2019-01-01"
end_date = "2020-12-31"
```

Perform the selection:

```{r}
my_selection <- eurobis |> 
  filter(aphiaidaccepted == aphia_ID,
         longitude > sel_longitude[1], 
         longitude < sel_longitude[2],
         latitude > sel_latitude[1], 
         latitude < sel_latitude[2],
         observationdate >= as.POSIXct(start_date),
         observationdate <= as.POSIXct(end_date)) |> 
  collect()
```

Inspect the selection, for herring you should have `137.152` occurrences between 2010-2020 in region of interest (lat(-20; 40), lon(30; 65)).

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Step 4: Visualize your selection on a map

Plot the data on a map using the `mapview` package. Notice that plotting might take some time, depending on the number of records in the dataset.

```{r}
mapview::mapview(my_selection$longitude, my_selection$latitude, crs = "epsg:4326")
```

# 2. ENVIRONMENTAL DATA

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Step 1: Source editoTools

Source this R script, it contains several useful functions to interact with the data lake.

```{r}
source("https://raw.githubusercontent.com/EDITO-Infra/R_tools/main/editoTools.R")
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Step 2: inspect EDITOSTAC

`EDITOSTAC` is a data frame from `editoTools` and contains a library of all available datasets.

Look at the data frame in your environment to see which datasets are available.

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Step 3: set search parameters

Following parameters are used:

-   `variable`: String (e.g. `thetao`, `so`, `zooc`, `phyc`)
-   `StacCatalogue`: use `EDITOSTAC` which is a data frame created in `editoTools`.
-   `lon_min`: int or float.
-   `lon_ma`: int or float.
-   `lat_min`: int or float.
-   `lat_max`: int or float.
-   `requestedTimeSteps`: can be one of `NA 86400000 21600000 3600000 10800000 900000 604800000`
-   `date`: single date Note:Just a date, doesn't work with multiple dates, would need to loop over it one-by-one
-   `select_layers`: NULL or int, if multiple layers available it will ask which one you want, can pre-select the first, second, ... layer

```{r}
variable = "thetao"
stacCatalogue = EDITOSTAC
lon_min = -12
lon_max = 10
lat_min = 48
lat_max = 62
requestedTimeSteps = 3600000
date = "2020-09-01"
```

### Step 4: request raster values

This function might give you some warnings, but it runs just fine.

```{r}
raster_example <- getRasterSlice(variable,
                                   stacCatalogue = EDITOSTAC,
                                   lon_min = lon_min,
                                   lon_max = lon_max,
                                   lat_min = lat_min,
                                   lat_max = lat_max,
                                   requestedTimeSteps = requestedTimeSteps,
                                   date = date,
                                   select_layers = NULL)
```

### Step 4: plot the raster as a map

```{r}
 # plot
mapview(raster_example)
```

# Part 3 TRAINING DATA

If we want to train a habitat suitability model, we would need the occurrences together with the environmental variables. For this, you can use the `enhanceDF` function. This function requires following columns: Latitude, Longitude and Time (YYYY-MM-DD)

```{r}
my_selection$Latitude = my_selection$latitude
my_selection$Longitude = my_selection$longitude
my_selection$Time = my_selection$observationdate
```

Next, we need to specify the required parameters as follows:

```{r}
parameters = list(
  "thetao"= c("par" = "thetao", "fun" = "mean", "buffer" = "1000"),
  "so"= c("par" = "so", "fun" = "mean", "buffer" = "1000")
  )
```

Run `enhanceDF`

NOTICE: this part includes command line interaction which is not enabled with the "run Current Chunk" play button. Instead, put your cursor before the first line and press cntrl + enter. If multiple data sources are available, the command line will ask your input in defining which layer it should use.

```{r}
df_occ_bg_env = enhanceDF(inputPoints = my_selection,
                          requestedParameters = parameters,
                          requestedTimeSteps = NA,
                          stacCatalogue = EDITOSTAC,
                          verbose="on")

```

# Discover more

`EditoTools.R` contains several other useful functions which are demonstrated in the tutorial "*Using EditoTools for predictive modelling of Atlantic herring larvae in the North Sea***".**

**\>\> Go ahead and explore this notebook :-)**
